\acknowledge{I thank my advisor, Mike Harms, for his excellent mentorship and his patience. He has pushed me to wrestle with new concepts and given me the opportunity to learn a diverse set of skills. I also thank the members of the Harms group for their 
constructive feedback, helpful conversations, and friendship throughout my time in the lab. I would especially like to thank Zach Sailer for his 
friendship and for always being willing to help me wrestle with new ideas. Thanks are also in order for the members of the Beer and Theory Society, who have devoted their time weekly, over the course of several years, to creating an excellent environment for learning math and physics that we all wish we'd done a better job of learning in college. In the same vein, I thank the members of the Quantitative Problem Solving and Research Communication Consortium for their devotion to helping peers solve challenging problems and representing the ideas of open science and collaboration. Thanks are in order for my long-time roommates Adam and Forrest as well as for the members of my trivia team. We've had a lot of good times during my years in Eugene. My friend Stacey Wagner has been very helpful to me over the years and I appreciate her willingness to get together and trade advice. The research in this dissertation was supported in part by a grant, R01GM117140, from the National Institutes of Health, and I was personally supported by NIH training grant T32 GM007759 for three years of my PhD. I thank my committee for their assistance throughout my PhD work and with the preparation of this document. Special thanks go to Jim Prell, who has been an excellent committee chair as well a collaborator. I have been inspired by his knowledge, enthusiasm, and strong distaste for false dichotomies. Furthermore, I thank Jim Prell and Alice Barkan for their assistance in applying for postdoctoral positions. I also thank Micah Donor, Shion An-Lim, and Susan Marqusee with whom I have collaborated. I thank Doug Turnbull and Maggie Weitzman in GC3F for their help with next-generation sequencing. I thank Carol Higginbotham at COCC and Ed LaChapelle at Bend Research, Inc. for encouraging me to pursue a scientific career very early in my life. I thank my undergraduate research advisor, Trevor Douglas, for helping me to develop a firm footing in biochemistry, molecular biology, and experimental design before starting graduate school. I would like to thank my good friend Ryan Russel Allen for his friendship, support, and steady stream of humorous correspondence throughout my undergraduate and graduate careers. Finally I would like to thank my family and my girlfriend for their constant support and love.}

